Dynamical properties of the Benguela Current as seen through a cold-spell algorithm.

### 2019-02-25 Write_up/Writing/dissertation.Rmd
Updated_commit

Complete datasets are often difficult to

@@

@@

@@

@@

# Chapter 2: Upwelling

## Introduction

@@

>>>>

Reliable datasets are often difficult to maintain and obtain. In the field of biological science research, it is often abiotic data products that are impaired thus leading to the use of satellite derived Sst products for the study of specific fields such as coastal and marine ecology as well as oceanography. These datasets however, may be inaccurate along the coast and local in situ collected data should be used when possible (Smit et al., 2013). Remotely sensed data products are obtained by several different organisations worldwide. Satellites launched into space detect the infra-red light emitted by the ocean which is then taken as a proxy from SST. Given the extensive distance at which these satellites record data, a certain level of coarseness is displayed within the dataset. This is demonstrated as the size of the pixels one sees in the final product.

Satellite SST products undergo a range of quality control and validation processes (e.g. Reynolds and Smith, 1994; Brown et al., 1999; Martin et al., 2012), making the data reliable and more useful. As previously mentioned, satellites may show a warm or cold bias along the coastal region but they are still useful for broader scale investigations (Smit et al., 2013, Smale and Wernberg, 2009; Castillo and Lima, 2010; Dufois and Rouault). Remotely-sensed SST products are obtained at different resolutions, and as a result do not show the same level of bias against in situ collected seawater temperature. Research done by Smit et al (2013) showed that MODIS represented the least amount of bias.

@@

### Event detection

As specified earlier, the extreme cold events occurring along the west coast of South Africa, which a central to this thesis will be detected, and correlated to upwelling events with the algorithm developed by Hobday et al. (2016). The events that recorde in situ data from SACTN would be compared against those detected in satellite derived SST. For this analyses the R package ‘RmarineHeatWaves’ and ‘heatwaveR’ were used to calculate the extreme events. In order to obtain the basis for these calculations, the temperature metrics must be aggregated into mean values based on the year in which they occurred. With these annual values it is possible to calculate long-term trends in extreme events.

## Analyses

The analyses conducted in this thesis varied according to the investigations performed. The one common thread uniting all of the analyses was that they were conducted in R-programming language (R Core Team, 2017). This provided reproducibility for all of the analyses and visualisations used.

@@

The primary aim of the first study was to make use of various remotely sensed SST data, in-situ collected coastal seawater temperature data and chlorophyll-a data. It was also useful to examine how the intensity and duration of upwelling varies per dataset as these datasets are collected at different resolutions, with the assumption that high resolution data may reveal a more accurate detection of the duration and intensity of upwelling events. Wind data used in these analyses were predicted to show clear patterns between weak and favourable upwelling winds and its impact on the intensity and duration of upwelling. Chlorophyll a data are then used to examine whether or not the intensity and duration of upwelling may lead to an increase or a decrease in the chlorophyll a concentration at the various sites. It was hypothesised that an increase in the intensity and duration of upwelling will show an increase in the abundance of chlorophyll a concentration. The testing of the hypothesis would be made possible with a series of analyses that allows one to select specific time periods from a range of datasets.

@@

The primary aim of the second study was to determine the occurrence of MCSs along the west coast of Southern Africa between the nearshore (<400m from the shore) and offshore. The most appropriate dataset for the measurement of the nearshore temperatures within South Africa is the South African Coastal Temperature Network (SACTN) dataset (Smit et al. 2013). With the knowledge obtained from study 1, it was now possible to investigate whether a more intense upwelling event may lead to an increase in  the intensity and duration of MCSs. It is decided that an investigation seeking to answer this, taken from historical records of the occerence of every thermal  event recorded along the west coast of South Africa for both in situ and remotely sensed seawater temperatures. The SACTN dataset and a total of five SST datasets could be used to investigate the occurrence of marine cold-spells (MCSs) with the methodology introduced by Hobday et al. 2016. This research also assists in the understanding of  local oceanographic patterns along the west coast of South Africa and the impact anthropogenically forced climate change may have for the recent changes in patterns and long-term trends.

## Research assumptions

The central assumption of this project was that remotely-sensed SST data and in situ seawater temperature data  may be compared. Remotely sensed SST data and in situ seawater temperature data were obtained at different resolutions and so slight changes may be found within the different datasets. An additional assumption within this research focused on the effect of upwelling on MCSs. Very little research has been done on the driving factors leading to MCSs and which processes may result in an increase in the intensity and duration of these events. For the purpose of this research  the influence of upwelling on MCSs was the primary focus.

@@

### Chapter 2 Upwelling

##Introduction

Upwelling events are known to have a large influence on the abundance, diversity and distribution of marine organisms and critically affect marine ecosystem processes. The changes in the intensity and duration of upwelling events have proven to cause large changes in ecosystem functioning [@wang]. Globally, there are four major eastern boundary upwelling systems (EBUS). Each of these systems are characterized as vast regions of coastal ocean occurring along the western shores of continents bordering the Pacific and Atlantic Oceans [@bakun ; @pauly]. Eastern boundary upwelling regions only cover 1% of the worlds ocean surface, but are among the most productive regions of the world. This is because these regions are influenced by upwelling events. Upwelling is defined as a process whereby cold, nutrient rich waters are pushed to the surface as a result of alongshore equatorial winds. The South African coastline, being approximately 3000km in length, is unique in that it is bordered by the Agulhas current along the east coast and the Benguela current, an eastern boundary upwelling system, along the west coast. These regions display noticeable differences in seawater temperatures in comparison to each other, primarily due to the influences of the neighbouring ocean currents (Bolton et al. 2004; Mead et al. 2013; Schlegel and Smit 2016). The Benguela Current is driven by the anticyclone high pressures systems and eastward moving cyclones which determine the boundaries of the Benguela upwelling region (Shannon 2006; Hutchings et al. 2009). The atmospheric temperatures along the coastline are largely influenced by the Benguela Current [@van], however, upwelling systems may be forced by various atmospheric conditions [@Lutjeharms; @roberts; @hutchings]. Regionally, the Benguela Current assists in transporting cold water northwards from the Southern Ocean to the coast (Lüning 1990; Lutjeharms, Cooper, and Roberts 2000; Hutchings et al. 2009; Schlegel et al. 2017).

Over the past three decades, climate change as a result of anthropogenic activities has established global concern and is shown to have negatively affected marine and terrestrial ecosystems. The consequences of these changes are also shown to have direct effects on the behaviour, survival and functioning of marine organisms, thus ultimately affecting species trophic levels and integration [@doney]. According to the Bakun hypothesis, an increase in greenhouse gases, will result in an increase in day time warming and night time cooling, this ultimately causes an increase in temperature gradients and this further results in a strong pressure gradient. Pressure gradients are known to affect wind patterns and wind patterns subsequently control the intensity and duration of upwelling events  [@hsieh; @mote; @lima].

Several research studies provide significant evidence suggesting an increase in ocean surface temperatures. These temperatures are agreed to have increased as a result of climate change  [@gruber]. This warming is shown to slowly move down into the water column thus increasing thermal stratification. This thermal stratification restricts the depth at which water is upwelled and thus limits the amount of nutrients brought up to the surface with a decrease in oxygen and pH levels [@chhak; @jacox2011; @jocox2016]. This process is thus strongly linked to primary productivity. Climate change impacting ocean processes has major ecological consequences (Bradley et al. 1999; Walther et al. 2002). For example, several regions show an increase in sea surface temperatures. This rise in temperatures largely influence species survival, reproduction and abundance. These changes are known to yield several economic consequences (Murawski 1993), thus, understanding the impact and cause of climate change is a challenging phenomenon. Similarly, climate change also largely influences coastal species such as corals, muscles and oysters. Given the importance of primary production, it was determined that the Benguela upwelling system, the eastern boundary upwelling system in South Africa, should be the focus of this paper.

The Benguela system is bounded by the equatorward and poleward ends by warm water regimes. These currents are shown to be forced by local wind stress fields off southwest Africa[@nelson ; @fennel]. This EBUS streches along the south-western coast of Africa from southern Angola to Cape Agulhas [@cole]. Water is coldest and most persistent off the Namibian coast at 25°S and weakest at 17°S [@Lutjeharms].  The upwelling period in the benguela north of 25°C  persist throughout the entire year but peaks during August. The Benguela upwelling region is split into northern and southern systems by a zone of intense perennial upwelling activity in the Luderitz region within the Namibian region [@shannon; @cole]. Several research studies have shown extreme warming and cooling periods occurring in the northern and southern benguela systems [@stander; @walker; @shannon1984; @boyd; @brainard]. Research studies done by Walter (1937) also show anomously warm periods which may have resulted as a lack of upwelling or poor upwelling events. During the periods of warming the equatorward wind stress over the northern region of the benguela system was stronger than usual; yet upwelling was suppressed by the southward movement of warm saline waters from the Angolan region. In the southern benguela region, continental shelf bathymetry and strong upwelling favourable winds are known to result in more intense upwelling events. Upwelling occuring along the south coastline of the of Southern Africa are particularly generated by local winds.

Global warming is increasing at an extremely high rate. This will result in an increase in greenhouse gases which in turn will impact the ocean and terrestrial life. The process of upwelling is one such factor that is predicted to change with an increase in global warming. As prreviously mentioned, changes in temperature results in changes in pressure gradients, these pressure gradients control wind patterns which ultimately controls the intensity and duration of upwelling. To assess the changes within the upwelling events, various datasets would be used. These are satellite sea surface temperatures (SST), in situ collected seawater temperatures, wind data as well as satellite obtained chlorophyll-a data.

@@

In order to compare abiotic variables such as wind action and direction along the South African coastline, large historical datasets for temperature and wind were analysed and accessed.

## Wind data

Wind action was an important variable in this study as it was hypothesised to exhibit a direct
influence on seawater temperatures (Sinnett and Feddersen 2014) and a direct influence on the intensity and duration of upwelling events; consequently, wind action was investigated for its impact on upwelling at specific sites along the west coast of South Africa. Wind data was obtained from the South African Weather Service (SAWS), and were provided at three-hour resolutions. Specific wind characteristics were measured, namely, wind direction (dirw) and wind speed (spw).  A resolution of 200m was modelled at both 7 and 15m isobaths.

## Temperature data sources

This research study lies within the borders of South Africa along the west coast in particular. The locations of each of the points used within this study can be seen in Fig.1. The South African Coastal Temperature Network was the primary dataset used to specify the various sites along the west coast of South Africa. This dataset consists of coastal seawater temperatures obtained from 129 sites along the entire South African coastline which is approximately 3000km long and is distinct in that it is bordered by the Benguela and Agulhas Currents (e.g.,Roberts 2005; Hutchings et al. 2009). These currents are known to largely influence the seawater temperature and marine life along the South African coastline. The SACTN has collected SST data along this unique coastline from as early as 1972 till present date, with contributions made from several organisations and govermental departments. As a result, temperature within this datasets were obtained by using different methods and instruments, such as hand held thermometers or Underwater temperature reorders( UTRs). The precision of the instruments used vary accordingly, with hand-held thermometers recording temperature at a precision of 0.5°C as opposed to the underwater temperature recorders which have an instrument precision of 0.001°C. The duration and extent of recordings per sites were uneven with the longest time series in the dataset being that of Gordons Bay which started on 13 September 1972 and concluded on 26 January 2017.  During the 1970s, a total of 11 time series began recording. A further 53 entries were added during the 1980s, 34 entries were added during the 1990s, and 18 entries were added during the 2000s. Recordings are still ongoing at many of these sites.

This study also made use of a range of remotely sensed temperature datasets. Satellites provide a system of large spatial and temporal data retrieval. Remotely sensed SST products are produced by a number of organisations across the globe. The AVHRR-only Optimally-Interpolated Sea Surface Temperature (OISST) was used to determine SST within the study region. The AVHRR datasets have been provided global SSTs for more than four decades (Reynolds and Smith, 1994; Pearce et al 2006). OISST is a global 1/4° gridded daily SST product that assimilates both remotely sensed and in situ sources of data to create a level-4 gap free product (Banzon et al., 2016). The Multi-scale Ultra-high Resolution (MUR) Sea Surface Temperature Analysis, the second dataset, is produced using satellite instruments with datasets spanning 1 June 2002 to present times (PO.DAAC, https://podaac.jpl.nasa.gov/). MUR provides SST data at a spatial resolution of 0.01° in longitude-latitude coordinates and is currently among the highest resolution SST datasets available. The CMC dataset constitutes the third dataset and is a version 3.0 Group for High Resolution Sea Surface Temperature (GHRSST) Level 4 dataset with a 10km resolution constructed by the Canadian Meteorological Center (CMC). The CMC dataset combines infrared satellite SST at numerous points in the time series from the AVHRR, the European Meteorological Operational-A (METOP-A) and Operational-B (METOP-B) platforms, and microwave SST data from the Advanced Microwave Scanning Radiometer 2 in conjunction with in situ observations of SST from ships and buoys from the ICOADS program.

## Chlorophyll-a data

In order to observe the patterns in the variation of the chlorophyll-a concentration with the intensity and duration of upwelling events,  MODIS-Aqua Chlorophyll a concentration, OCI Algorithm, 8-day composite were used. The National Aeronautics and Space Administration’s (NASA) Moderate Resolution imaging Spectroradiometer (MODIS) Aqua satellite is an example of a satellite currently collecting the reflected wavelength energy from the earth. The MODIS Aqua satellite records wavelength data down to the 1% photosynthetically available depth (0.1 PAR depth) of the water column, a vital component in the estimation of chlorophyll (NASA.GOV, http://oceancolor.gsfc.nasa.gov). Because chlorophyll-a concentration retrieval is only possible over cloud-free areas the daily data offered may have large data gaps due to cloud cover. The quality of the chlorophyll-a concentration  estimates depends strongly on the quality of the cloud mask. The NASA Aqua satellite contains the MODIS sensor that is used for detecting chlorophyll-a concentrations (ocean color) in the world's oceans, among other applications (NASA.GOV, http://oceancolor.gsfc.nasa.gov). The Level 3 standard mapped image (SMI) chlorophyll-a dataset has a monthly temporal resolution and 4.6 km spatial resolution. The SMI dataset is an image representation of binned MODIS data.

## Site selection

In order to compare the impact of wind action on the intensity and duration of upwelling, the appropriate sites along the west coast of South Africa (i.e. sites influenced by the Benguela current) were selected. Since temperature data were not evenly recorded for each of the sites, Firstly the full dataset was narrowed to only those sites that could be adequately compared. To do this, the filter () function in R was used. This function selects all the sites with a time series of greater than 10 years. Once the sites were minimised, one years worth of data was selected from the SACTN data between the years 2014 – 2015.  This yield a total of  seven sites along the west coast of South Africa. Hereafter, each of the remotely sensed SST datasets obtained were matched with the SACTN dataset. This was done for two reasons, to find the exact same points at which temperature was collected for the SACTN sites and secondly, to confirm that the dates and time that temperature were collected correspond accordingly.
